Article Title

### 2014-01-22 plos.tex
fixed a buncha things in the manuscript

\begin{FPfigure}
    \begin{center}
        \includegraphics{./figures/Figure02.pdf}
    \end{center}
    \caption{{\bf Regularized estimators whose structure matches the true structure in the data are more efficient.}
        {\bf Row 1.} Graphical representations of the target estimates used of the four respective regularized covariance matrix estimators.  Recorded neurons are denotes as green spheres and latent units as light-shaded spheres.  The edges between them denote non-zero partial correlations.
        {\bf Row 1, A}.  In estimator $C_{\sf diag}$, the target estimate is a diagonal matrix, which describes systems that lack linear associations. 
        {\bf  Row 1, B.} In estimator $C_{\sf factor}$, the target estimate is a factor model (low-rank matrix plus a diagonal matrix), favoring systems in which correlations arise due to common input from latent units. 
        {\bf  Row 1, C}. In estimator $C_{\sf sparse}$, the covariance matrix is approximated as the inverse of a sparse matrix. This approximation describes systems in which correlations arise from linear associations between the observed units.
        {\bf  Row 1, D}.  In estimator $C_{\sf sparse+latent}$, the covariance matrix is approximated as the inverse of the sum of a sparse matrix and a low-rank matrix. This approximation describes a model wherein correlations arise due to linear associations between the recorded cells \emph{and} with several latent units.
\\
{\bf Row 2:} Examples of $50\times 50$ correlation matrices corresponding to each structure: {\bf A.} the diagonal correlation matrix, {\bf B.} a factor model with two latent units, {\bf C.}  a correlation matrix with 85\%  off-diagonal zeros, and {\bf  D.} a correlation matrix whose inverse is the sum of a rank-1 (\ie one latent unit) matrix and a sparse matrix with 89\% off-diagonal zeros. 
\\
{\bf Row 3:} Sample correlation matrices calculated from samples of size $n=1000$ drawn from simulated random processes with respective correlation matrices from row 2.  The structure of the sample correlation matrix is difficult to discern by eye.
\\
{\bf Row 4:} Correlation matrix estimates computed by estimators with matching structure from the same data as the sample correlation matrices in Row 3. The regularized estimates are closer to the truth than the sample correlation matrices.
\\
{\bf Row 5:} Excess losses (Eq.~\ref{eq:excess-loss}) attained by the five estimators as a function of sample size. The error bars indicate the standard error of the mean based on 30 samples.  Estimators with the matching structure converged to zero faster than the other estimators. Some exceptions occur when the sample is not large enough to reveal the true structure; in these cases estimators with simpler structures can be more efficient.
\\
{\bf Row 6:} Relative cross-validation losses (Eq.~\ref{eq:rel-cv-loss}) attained the five estimators with respect to the matching estimator. Error bars indicate the standard error of the mean based on 30 samples.
    }
    \label{fig:02}
\end{FPfigure}

@@

\paragraph{Simulation}
To demonstrate how the selection between regularized estimators allows discerning the structure of correlations, we constructed four $50\times 50$ covariance matrices with the structures imposed by the four respective regularized estimators (\figref{02}{\,Row 2, A--D}).  See the Methods section for details of how these matrices were generated.  We used these covariance matrices as the ground truth in multivariate normal distributions with zero means and drew samples of various sizes. Even with relatively large sample sizes (\eg $n=1000$), the sample correlation matrices were contaminated by sampling noise and their underlying structure was difficult to discern (\figref{02}{\,Row 3}).

An evaluation of a covariance matrix estimators requires a \emph{loss function} $\loss{C,\Sigma}$ to quantify the discrepancy between the estimate, $C$, and the truth, $\Sigma$.  The loss function must attain its minimum when $C=\Sigma$.

@@

We drew 30 independent samples from each model of size $n=250$, 500, 1000, 2000, and 4000 and measured the excess loss produced by each of the five estimators.  The hyperparameters of the regularized estimators were optimized by nested cross-validation using only the data in the sample.  All the regularized estimators produced better estimates (lower excess loss) than the sample covariance matrix.  However, estimators whose structure matched the true model typically outperformed the other estimators (\figref{02}{\,row 5}).  There were two exceptions: First, for some smaller samples, the data were insufficient to reveal the true correlation structure and the estimators with simpler structures, $C_{\sf diag}$ and $C_{\sf factor}$ often outperformed the estimators with matching structure (\figref{02}{\,Row 5, C and D}).  Second, $C_{\sf sparse+latent}$ often performed equally well to $C_{\sf sparse}$ on the model without latent units. This can be explained by the fact that $C_{\sf sparse+latent}$ correctly inferred the right number of latent units, zero.

@@

In empirical data, the ground truth, $\Sigma$, is not accessible and excess loss cannot be computed directly. Instead, the loss function can be estimated solely from the data through \emph{validation}.  In validation, an independent \emph{testing sample} is set aside to compute an additional sample covariance estimate $C_{\sf 0}^\prime$ to validate the estimate $C$ computed from the main, \emph{training sample}.  Then \emph{validation loss} $\loss{C,C_{\sf 0}^\prime}$ can substitute the loss function $\loss{C,\Sigma}$.

@@

In this case, the cross-validation loss is an unbiased estimate of the loss function:
\begin{equation*}
    \E[C_{\sf 0}^\prime]{\loss{C,C_{\sf 0}^\prime}}=\loss{C,\E[C_{\sf 0}^\prime]{C_{\sf 0}^\prime}}=\loss{C,\Sigma}
\end{equation*}

@@

Acquiring a separate testing sample is not practical. Instead, $K$-fold cross-validation can be used: The sample is divided into $K$ subsets of approximately equal size ($K=10$ in this study).  Then each subset is used as the validation sample with the other $K-1$ serving as the training sample. The validation losses from each of such \sq{folds} are averaged to produce the \emph{cross-validation loss} or \emph{CV-loss} for short.  Let $C_{\sf 0}^{\{k\}}$ denote the sample covariance matrix computed from the $k^{th}$ subset and $C^{\{\setminus k\}}$ denote the results of estimator $C$ trained on the remaining $K-1$ subsets. Then cross-validation loss for estimator $C$ is
\begin{equation}\label{eq:cv-loss}
    \ell_C=\frac 1 K \sum\limits_{k=1}^K \loss{C^{\{\setminus k\}},C_{\sf 0}^{\{k\}}}
\end{equation}
In this formulation, $\ell_C$ can be recognized as the cross-validated Gaussian log-likelihood (up to a constant offset and multiplier). However, the present framework is more general, allowing for other loss functions to be used.

Since we only need to compare estimators to each other, we are only interested in the \emph{relative} CV loss of estimator $C$ with respect to reference estimator $C_{\sf ref}$:
\begin{equation}\label{eq:rel-cv-loss}
    \Delta\ell_{C,C_{\sf ref}} = \ell_{C}-\ell_{C_{\sf ref}}
\end{equation}

To examine how well the relative CV-losses reproduced the relationships between the excess losses, we evaluated the relative CV-loss with respect to the estimator that matched the structure of the data-generating model. Indeed, CV-losses accurately reproduced the relationships between the excess losses of the estimators, with somewhat greater variability (\figref{02}{Row 6}).

This simulations study demonstrated that, with sufficiently large sample sizes, imposing the correct type of structure on the estimate, led to greater estimation improvement than imposing incorrect structures

\paragraph{Covariance estimation in neural data}
We recorded the calcium activity of dense populations of neurons in the supragranular layers in primary visual cortex of anesthetized mice using fast random-access 3D scanning two-photon microscopy during visual stimulation (\figref{03}{A--B}) \cite{Reddy:2005, Katona:2012, Cotton:2013}. This technique allowed fast sampling (100--150 Hz) from large numbers (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4 (\figref{03}{C}).  The firing rates of neurons were inferred using sparse nonnegative deconvolution \cite{Vogelstein:2010} to produce (\figref{03}{C and D}). Only cells that produced detectable calcium activity were included in the analysis (For details, see Methods).  First, full-field gratings moving in 16 directions were presented to determine the orientation tuning of the recorded cells (\figref{03}{E}). To estimate the noise correlation matrix, we presented between 100 and 300 repetitions of full-field drifting gratings at each direction of motion.   Each grating lasted 1 second and was followed by a 1-second blank.  The average stimulus response was subtracted from each trial; the traces were then binned into 150 ms intervals for the estimation of the covariance matrix.   The sample correlation coefficients were largely positive and low (0.01--0.03 in most images sites) (\figref{03}{F and G}).

\begin{figure}    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,center},capbesidewidth=8.3cm}}]{figure}[\FBwidth]
    {\caption{{\bf Acquisition of neural signals for the estimation of noise correlations.}
    Visual stimuli comprising full-field drifting gratings interleaved with blank screens ({\bf A}) presented during two-photon recordings of somatic calcium signals using fast 3D random-access microscopy ({\bf B}). 
    {\bf C--G.} Calcium activity data from an example site.
    {\bf C.} Representative calcium signals from eight cells out of 298 cells downsampled to 20 Hz. The inferred firing rate binned in 150 ms intervals are indicated by red ticks below each trace.
    {\bf D.} The raster plot of the inferred firing rates, binned in 150 ms intervals, from the entire population from the first (left) and last (right) minute of the entire recording.  The traces from {\bf C} are highlighted in red.
    {\bf E.} The spatial arrangement and orientation tuning of the 298 cells from the imaged site.
    {\bf F.} The noise correlation matrix of the activity of the neural population. 
    {\bf G.} Histogram of noise correlation coefficients. The red line indicates the mean.
} \label{fig:03}}
    {\includegraphics[width=8.3cm]{figures/Figure03.pdf}}
\end{figure}

In these dense local populations direct interactions between cells are likely to contribute substantially to the patterns of population.  We therefore hypothesized that covariance matrix estimates that explicitly model the partial correlation between pairs of neurons ($C_{\sf sparse}$ and $C_{\sf sparse+latent}$ would have a performance advantage over those that do not.  However, the observed neurons must also be strongly influenced by global activity fluctuations and by unobserved common inputs, giving an advantage to estimators that explicitly model common fluctuations of the entire population $C_{\sf factor}$ and $C_{\sf sparse+latent}$.  If both types of effects are significant, then $C_{\sf sparse+latent}$ should outperform the other estimators.

To test this hypothesis, we estimated the relative CV loss (Eq.~\ref{eq:rel-cv-loss}) of estimators  $C_{\sf 0}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ with respect to $C_{\sf sparse+latent}$.  The hyperparameters of each estimator were optimized by nested cross-validation (See Methods for details). Indeed, the median relative CV loss was significantly above zero (Fig.~\ref{fig:04}), indicating that $C_{\sf sparse+latent}$ produced estimates that were consistently closest to the true covariance matrix.
\begin{figure}[!ht]    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,center},capbesidewidth=8.3cm}}]{figure}[\FBwidth]
    {\caption{{\bf The sparse+latent estimator $C_{\sf sparse+latent}$ outperforms the other estimators on neural data.}
    {\bf A--D.} Histograms of average cross-validation loss differences of the respective estimators $C_{\sf 0}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ from $C_{\sf sparse+latent}$. 
    The histograms are based on 31 imaged sites in 15 mice. 
    All medians (red dashed lines) were significantly greater than zero, indicating the dominance of $C_{\sf sparse+latent}$ over the other estimators. 
    The arrow heads indicate the results for the site shown in Fig.~\ref{fig:03} and Fig.~\ref{fig:05}.
\\{\bf A.} $C_{\sf sparse+latent}$ outperforms $C_{\sf 0}$: median improvement 0.16 nats/neuron, $p=3.9\times 10^{-5}$.
\\{\bf B.} $C_{\sf sparse+latent}$ outperforms $C_{\sf diag}$: median improvement 0.0032 nats/neuron, $p=2.2\times 10^{-3}$.
\\{\bf C.} $C_{\sf sparse+latent}$ outperforms $C_{\sf factor}$: median improvement 0.16 nats/neuron, $p=6.9\times 10^{-5}$.
\\{\bf D.} $C_{\sf sparse+latent}$ outperforms $C_{\sf sparse}$: median improvement 0.0016 nats/neuron, $p=5.5\times 10^{-6}$.
} \label{fig:04}}
    {\includegraphics{./figures/Figure04.pdf}}
\end{figure}

@@

Evaluation with respect to the quadratic loss function,
\begin{equation}\label{eq:quadratic}
\loss{C,C_{\sf 0}^\prime}=\frac 1 {p^2}\Tr(C^{-1}C_{\sf 0}^\prime-I)^2,
\end{equation}
produced the same relationship between the estimators as the normal loss (Fig.~S\ref{supp:02}) suggesting that the results of this study do not depend on the assumption of gaussianity and reflect real regularities in the data.

\paragraph{Correlation structure and circuit architecture}
Because of its superior estimation performance, $C_{\sf sparse+latent}$ likely provides a more physiologically relevant representation of the functional connectivity than the other estimators.  The estimator segregates two sources of correlations: linear dependencies between pairs of neurons (sparse component) suggestive of direct interactions and common fluctuations across the entire population (low-rank) component suggestive of common inputs, cell assembly activation, network fluctuations or other forms of synchronized activations.

We examined the structure of the $C_{\sf sparse+latent}$ estimate in each imaged site. The regularized estimate of the correlation matrix appeared very similar to the sample correlation matrix (\figref{05}{A and D}) but the corresponding partial correlation matrices differed substantially. The partial correlation matrix (\figref{05}{B and E})  and a low-rank component (\figref{05}{C}).  While sample correlations were mostly positive, the sparse partial correlations had a high fraction of negative interactions.  In the example site (Fig.~\ref{fig:03} and Fig.~\ref{fig:05})),  the sparse component had 82.2\% sparsity (or, conversely, 17.8\% connectivity) with average node degree of 52.8 (\figref{05}{G}).  The estimate also inferred 15 latent units contributing to the low-rank component of the partial correlation matrix.

\begin{FPfigure}
    \begin{center}
        \includegraphics[width=17.35cm]{./figures/Figure05.pdf}
    \end{center}
    \caption{{\bf Example of low-dimensional correlation structure revealed by the sparse+low-rank estimator.}
    {\bf A.} The regularized estimate of the correlation matrix (top-right) closely approximates the sample correlation matrix (bottom left). 
    This close approximation is  also demonstrated by the scatter plot of the correlation coefficients produced by the two estimates ({\bf D}). 
    However, the partial correlation matrices from the two estimates show more pronounced differences ({\bf B} and {\bf E}). 
    {\bf C.} The partial correlation matrix of the regularized estimate is decomposed into a sparse component with 82.2\% off-diagonal zeros (bottom-left) and low-rank component of rank 15 (top-right).
    {\bf F.} The sparse component of the regularized partial correlation matrix had little resemblance to the sample correlations. The gray interval indicates the range of correlations containing 82.2\% of cells pairs, equal to the fraction of zeros in the sparse partial correlation matrix. This interval contained 58.3\% of the partial correlations. 
    {\bf G.} A graphical depiction of the positive (green) and negative (magenta) partial correlations as edges between observed neurons. The line density is proportional to the magnitude of the correlation.
    {\bf H.} A subset of neurons from the center of the cluster shown in {\bf G} showing the regularized partial correlations.
    {\bf I.} The same subset with sample correlations thresholded to match the sparsity of the regularized interactions.
}
\label{fig:05}
\end{FPfigure}

We compared the graphs of partial correlations revealed by $C_{\sf sparse+latent}$ to the graph obtained by thresholding the sample correlations.   Thresholded correlations have been used in several studies as to isolate the likely pairwise interactions with subsequent graph theoretical analysis \cite{ Malmersjo:2013}.  We set the threshold to match the connectivity of the $C_{\sf sparse+latent}$ graph.  The networks revealed by the two methods differed substantially. In the example site with 17.8\% connectivity, only 41.7\% of cell pairs connected in the $C_{\sf sparse+latent}$ graph were also connected in the thresholded correlation graph (\figref{05}{F, H, and I}).   In particular, the thresholded sample correlations were predominantly positive whereas 43.3\% of the interactions in $C_{\sf sparse+latent}$ were negative.  Most of the negative interactions corresponded to low sample correlations (\figref{05}{F}).  Thus the $C_{\sf sparse+latent}$ estimates reveal negative interactions that account for low correlations where high correlations are predicted from all the other interactions.

The average partial correlations revealed by $C_{\sf sparse+latent}$ estimates in all 31 sites were about 5 times lower than the sample correlations and less variable across sites (\figref{06}{A}): the coefficient of variation of the average sample correlations across all sites was 0.41 whereas that of the average partial correlations was only 0.35.  The average node degree of the sparse component of the partial correlations and the number of inferred latent units varied widely between sites but generally increased with recorded population size (\figref{06}{B and C}). However, there was an inverse relationship between the number of latent units and the average node degree (\figref{06}{D}). Several sites, even with relatively large population sizes, had fairly few pairwise interactions and were dominated by latent units.  These differences have multiple explanations, including differences in brain states and recording quality (neuropil contamination, motion).

@@

We also examined the relationship of the correlation structure to the differences in orientation preference and to the geometric distances between the cell pairs (Fig.\;\ref{fig:07}).  The partial correlations of the $C_{\sf sparse+latent}$ estimates fell more rapidly with the difference in preferred orientation (\figref{07}{A}). Similarly, the regularized partial correlations decreased more rapidly than sample correlations with geometric distance laterally  (\figref{07}{B}) and in depth (\figref{07}{B}).

@@

Furthermore, the positive and negative interactions in the $C_{\sf sparse+latent}$ estimator were organized differently: The positive interactions fell rapidly with differences in the preferred orientation (\figref{07}{D}), lateral displacements (\figref{07}{E}), and displacement in depth (\figref{07}{F}) while the negative interactions were much less selective (\figref{07}{D--F}).

>>>>

\begin{FPfigure}
    \begin{center}
        \includegraphics{./figures/Figure02.pdf}
    \end{center}
    \caption{{\bf Regularized estimators whose structure matches the true structure in the data are more efficient.}
        {\bf Row 1.} Graphical representations of the target estimates used of the four respective regularized covariance matrix estimators.  Recorded neurons are denotes as green spheres and latent units as light-shaded spheres.  The edges between them denote non-zero partial correlations.
        {\bf Row 1, A}.  In estimator $C_{\sf diag}$, the target estimate is a diagonal matrix, which describes systems that lack linear associations. 
        {\bf  Row 1, B.} In estimator $C_{\sf factor}$, the target estimate is a factor model (low-rank matrix plus a diagonal matrix), favoring systems in which correlations arise due to common input from latent units. 
        {\bf  Row 1, C}. In estimator $C_{\sf sparse}$, the covariance matrix is approximated as the inverse of a sparse matrix. This approximation describes systems in which correlations arise from linear associations between the observed units.
        {\bf  Row 1, D}.  In estimator $C_{\sf sparse+latent}$, the covariance matrix is approximated as the inverse of the sum of a sparse matrix and a low-rank matrix. This approximation describes a model wherein correlations arise due to linear associations between the recorded cells \emph{and} with several latent units.
\\
{\bf Row 2:} Examples of $50\times 50$ correlation matrices corresponding to each structure: {\bf A.} the diagonal correlation matrix, {\bf B.} a factor model with two latent units, {\bf C.}  a correlation matrix with 85\%  off-diagonal zeros, and {\bf  D.} a correlation matrix whose inverse is the sum of a rank-1 (\ie one latent unit) matrix and a sparse matrix with 89\% off-diagonal zeros. 
\\
{\bf Row 3:} Sample correlation matrices calculated from samples of size $n=1000$ drawn from simulated random processes with respective correlation matrices from row 2.  The structure of the sample correlation matrix is difficult to discern.
\\
{\bf Row 4:} Correlation matrix estimates computed by estimators with matching structure from the same data as the sample correlation matrices in Row 3. The regularized estimates are closer to the truth than the sample correlation matrices.
\\
{\bf Row 5:} Excess losses (Eq.~\ref{eq:excess-loss}) attained by the five estimators as a function of sample size. The error bars indicate the standard error of the mean based on 30 samples.  Estimators with the matching structure converged to zero faster than the other estimators. Some exceptions occur when the sample is not large enough to reveal the true structure; in these cases estimators with simpler structures can be more efficient.
\\
{\bf Row 6:} Relative cross-validation losses (Eq.~\ref{eq:rel-cv-loss}) attained the five estimators with respect to the matching estimator. Error bars indicate the standard error of the mean based on 30 samples.
    }
    \label{fig:02}
\end{FPfigure}

@@

\paragraph{Simulation}
To demonstrate how the selection between regularized estimators allows discerning the structure of correlations, we constructed four $50\times 50$ covariance matrices with the structures imposed by the four respective regularized estimators (\figref{02}{\,Row 2, A--D}).  See the Methods section for details of how these matrices were generated.  We used these covariance matrices as the ground truth in multivariate normal distributions with zero means and drew samples of various sizes. Even with relatively large sample sizes (\eg $n=1000$), the sample correlation matrices were contaminated by sampling noise and their underlying structures were difficult to discern (\figref{02}{\,Row 3}).

The evaluation of a covariance matrix estimator $C$ is performed with respect to a \emph{loss function} $\loss{C,\Sigma}$ to quantify the discrepancy between the estimate and the truth, $\Sigma$.  The loss function must attain its minimum when $C=\Sigma$.

@@

We drew 30 independent samples of size $n=250$, 500, 1000, 2000, and 4000 from model and computed the excess loss resulting from the five estimators.  The hyperparameters of the regularized estimators were optimized by nested cross-validation using only the data in the sample.  All the regularized estimators produced better estimates (lower excess loss) than the sample covariance matrix.  However, estimators whose structure matched the true model typically outperformed the other estimators (\figref{02}{\,row 5}).  There were two exceptions: First, for some smaller samples, the data were insufficient to reveal the true correlation structure and the estimators with simpler structures, $C_{\sf diag}$ and $C_{\sf factor}$, often outperformed the estimators with the matching structure (\figref{02}{\,Row 5, C and D}).  Second, $C_{\sf sparse+latent}$ often performed equally well to $C_{\sf sparse}$ on the data from models without latent units because it reliably inferred the right number of latent units, zero.

@@

In empirical data, the ground truth, $\Sigma$, is not accessible and excess loss cannot be computed directly. Instead, the loss function can be estimated solely from the data through \emph{validation}.  In validation, an independent \emph{testing sample} is set aside to compute an additional sample covariance estimate $C_{\sf 0}^\prime$ to validate the estimate $C$ computed from the  \emph{training sample}.  Then \emph{validation loss} $\loss{C,C_{\sf 0}^\prime}$ can substitute the loss function $\loss{C,\Sigma}$.

@@

In this case, the validation loss is an unbiased estimate of the loss function:
\begin{equation*}
    \E[C_{\sf 0}^\prime]{\loss{C,C_{\sf 0}^\prime}}=\loss{C,\E[C_{\sf 0}^\prime]{C_{\sf 0}^\prime}}=\loss{C,\Sigma}
\end{equation*}

@@

Acquiring a separate testing sample is not practical. Instead, $K$-fold cross-validation can be used: The sample is divided into $K$ subsets of approximately equal size ($K=10$ in this study).  Then each subset is used as the validation sample with the other $K-1$ serving as the training sample. The validation losses from each of such \sq{folds} are averaged to produce the \emph{cross-validation loss} or \emph{CV-loss} for short.  Let $C_{\sf 0}^{\{k\}}$ denote the sample covariance matrix computed from the $k^{th}$ subset and $C^{\{\setminus k\}}$ denote the results of estimator $C$ trained on the remaining $K-1$ subsets. Then the cross-validation loss for estimator $C$ is
\begin{equation}\label{eq:cv-loss}
    \ell_C=\frac 1 K \sum\limits_{k=1}^K \loss{C^{\{\setminus k\}},C_{\sf 0}^{\{k\}}}
\end{equation}
In this formulation, $\ell_C$ can be recognized as the cross-validated Gaussian log-likelihood (up to a constant offset and multiplier). However, the present framework is more general, allowing for other loss functions to be used.

Since we only need to compare estimators to each other, we are only interested in the \emph{relative} CV loss of estimator $C$ with respect to reference estimator $C_{\sf ref}$:
\begin{equation}\label{eq:rel-cv-loss}
    \Delta\ell_{C,C_{\sf ref}} = \ell_{C}-\ell_{C_{\sf ref}}
\end{equation}
The relative cross-valiation loss was used for the main findings in this study.

To examine how well the relative CV-losses reproduced the relationships between the excess losses, we evaluated the relative CV-loss with respect to the estimator that matched the true structure of the model model. Indeed, CV-losses accurately reproduced the relationships between the excess losses of the estimators, with somewhat greater variability (\figref{02}{Row 6}).

This simulation study demonstrated that, with sufficiently large sample sizes, imposing the correct type of structure on the estimate led to greater estimation improvement than imposing other structures.

\paragraph{Covariance estimation in neural data}
We recorded the calcium activity of dense populations of neurons in the supragranular layers in primary visual cortex of fentanyl-anesthetized mice using fast random-access 3D scanning two-photon microscopy during visual stimulation (\figref{03}{A--B}) \cite{Reddy:2005, Katona:2012, Cotton:2013}. This technique allowed fast sampling (100--150 Hz) from large numbers (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4 (\figref{03}{C}).  The firing rates were inferred using sparse nonnegative deconvolution \cite{Vogelstein:2010} to produce (\figref{03}{C and D}). Only cells that produced detectable calcium activity were included in the analysis (For details, see Methods).  First, 30 repetitions of full-field drifting gratings of 16 direction were presented in random order for 500 ms without blanks.  This stimulus was used to compute the orientation tuning of the recorded cells (\figref{03}{E}). To estimate the noise correlation matrix, we presented only 2 or 5 distinct directions with between 100 and 300 repetitions of each direction. In this stimulus, each grating lasted 1 second and was followed by a 1-second blank.  The average stimulus response was subtracted from each trial. The traces were then binned into 150 ms intervals for the estimation of the covariance matrix.   The sample correlation coefficients were largely positive and low (0.01--0.03 in most images sites) (\figref{03}{F and G}).

\begin{figure}    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,center},capbesidewidth=8.3cm}}]{figure}[\FBwidth]
    {\caption{{\bf Acquisition of neural signals for the estimation of noise correlations.}
    Visual stimuli comprising full-field drifting gratings interleaved with blank screens ({\bf A}) presented during two-photon recordings of somatic calcium signals using fast 3D random-access microscopy ({\bf B}). 
    {\bf C--G.} Calcium activity data from an example site.
    {\bf C.} Representative calcium signals from eight cells out of 298 cells downsampled to 20 Hz. The inferred firing rate binned in 150 ms intervals are indicated by red ticks below each trace.
    {\bf D.} The raster plot of the inferred firing rates, binned in 150 ms intervals, from the entire population from the first (left) and last (right) minute of the entire recording.  The traces from {\bf C} are highlighted in red.
    {\bf E.} The spatial arrangement and orientation tuning of the 298 cells from the imaged site. The cells' colors indicate their orienation preferences. The gray cells were not significantly tuned to orientation.
    {\bf F.} The noise correlation matrix of the activity of the neural population. 
    {\bf G.} Histogram of noise correlation coefficients. The red line indicates the mean.
} \label{fig:03}}
    {\includegraphics[width=8.3cm]{figures/Figure03.pdf}}
\end{figure}

In these dense populations, direct interactions between cells are likely to influence the patterns of population activity.  We therefore hypothesized that covariance matrix estimates that explicitly model the partial correlation between pairs of neurons ($C_{\sf sparse}$ and $C_{\sf sparse+latent}$) would have a performance advantage.  However, the observed neurons must also be strongly influenced by global activity fluctuations and by unobserved common inputs, to the advantage of estimators that explicitly model common fluctuations of the entire population: $C_{\sf factor}$ and $C_{\sf sparse+latent}$.  If both types of effects are significant, then $C_{\sf sparse+latent}$ should outperform the other estimators.

To test this hypothesis, we computed the relative CV losses (Eq.~\ref{eq:rel-cv-loss}) of estimators  $C_{\sf 0}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ with respect to $C_{\sf sparse+latent}$ in 31 imaged sites in 15 mice.  The hyperparameters of each estimator were optimized by nested cross-validation (See Methods for details). Indeed, the median relative CV loss was significantly above zero (Fig.~\ref{fig:04}), indicating that $C_{\sf sparse+latent}$ produced estimates that were consistently closest to the true covariance matrix.
\begin{figure}[!ht]    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,center},capbesidewidth=8.3cm}}]{figure}[\FBwidth]
    {\caption{{\bf Estimator $C_{\sf sparse+latent}$ is more efficient than the other estimators on neural data.}
    {\bf A--D.} Histograms of the relative cross-validation loss of estimators $C_{\sf 0}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ with respect to $C_{\sf sparse+latent}$.
    The histograms are based on 31 imaged sites in 15 mice. 
    All medians (red dashed lines) were significantly greater than zero, indicating the dominance of $C_{\sf sparse+latent}$ over the other estimators. 
    The arrow heads indicate the results for the site shown in Fig.~\ref{fig:03} and Fig.~\ref{fig:05}.
\\{\bf A.} $C_{\sf sparse+latent}$ outperforms $C_{\sf 0}$: median improvement 0.16 nats/neuron, $p=3.9\times 10^{-5}$.
\\{\bf B.} $C_{\sf sparse+latent}$ outperforms $C_{\sf diag}$: median improvement 0.0032 nats/neuron, $p=2.2\times 10^{-3}$.
\\{\bf C.} $C_{\sf sparse+latent}$ outperforms $C_{\sf factor}$: median improvement 0.16 nats/neuron, $p=6.9\times 10^{-5}$.
\\{\bf D.} $C_{\sf sparse+latent}$ outperforms $C_{\sf sparse}$: median improvement 0.0016 nats/neuron, $p=5.5\times 10^{-6}$.
} \label{fig:04}}
    {\includegraphics{./figures/Figure04.pdf}}
\end{figure}

@@

The same evaluation based on the quadratic loss function,
\begin{equation}\label{eq:quadratic}
\loss{C,C_{\sf 0}^\prime}=\frac 1 {p^2}\Tr(C^{-1}C_{\sf 0}^\prime-I)^2,
\end{equation}
instead of the normal loss function reproduced the same relationship between the estimators (Fig.~S\ref{supp:02}). This suggests that the results of this study are robust to the choice of the loss function and do not depend on the assumption of gaussianity.

\paragraph{Correlation structure and circuit architecture}
Because of its superior estimation performance, $C_{\sf sparse+latent}$ likely provides a more physiologically relevant representation of the functional connectivity than the other estimators.  The estimator segregates two sources of correlations: linear dependencies between pairs of neurons (sparse component) suggestive of direct interactions and common fluctuations across the entire population (low-rank component) component suggestive of common inputs, cell assembly activation, network fluctuations or other forms of synchronized activations.

We examined the composition of the $C_{\sf sparse+latent}$ estimates in each imaged site. The regularized estimates appeared similar to the sample correlation matrix (\figref{05}{A and D}) but the corresponding partial correlation matrices differed substantially (\figref{05}{B and E}). The partial correlation matrix of the regularized estimate was decomposed into a sparse component and a low-rank component (\figref{05}{C}).  While sample correlations were mostly positive, the sparse partial correlations had a high fraction of negative interactions.  In the example site (Fig.~\ref{fig:03} and Fig.~\ref{fig:05}),  the sparse component had 82.2\% sparsity (or, conversely, 17.8\% connectivity) with average node degree of 52.8 (\figref{05}{G}).  The estimate also inferred 15 latent units contributing to the low-rank component of the partial correlation matrix.

\begin{FPfigure}
    \begin{center}
        \includegraphics[width=17.35cm]{./figures/Figure05.pdf}
    \end{center}
    \caption{{\bf Example of the structure revealed by $C_{\sf sparse+latent}$ the sparse+low-rank estimator.}
    {\bf A and D.} The regularized estimate $C_{\sf sparse+latent}$ closely approximates the sample correlation matrix $C_{\sf 0}$. 
    {\bf B and E.} However, the partial correlation matrices from the two estimates differ substantially.
    {\bf C.} The partial correlation matrix of the regularized estimate is decomposed into a sparse component with 82.2\% off-diagonal zeros (bottom-left) and low-rank component of rank 15 (top-right).
    {\bf F.} The sparse component of the regularized partial correlation matrix had little resemblance to the sample correlations: the gray interval indicates the range of correlations containing 82.2\% of cells pairs, equal to the fraction of zeros in the sparse partial correlation matrix. The significant correlations were outside this interval. Yet 58.3\% of the interactions inferred by $C_{\sf sparse+latent}$  linked pairs of neurons whose correlation was below the threshold. Only the remaning 41.7\% of the interactions overlapped with sample correlations above the theshold.
    {ebf G.} A graphical depiction of the positive (green) and negative (magenta) partial correlations as edges between observed neurons. The line density is proportional to the magnitude of the correlation.
    {\bf H.} A subset of neurons from the center of the cluster shown in {\bf G} showing the regularized partial correlations.
    {\bf I.} The same subset with sample correlations thresholded to match the sparsity of the regularized interactions.
}
\label{fig:05}
\end{FPfigure}

We compared the graphs of partial correlations revealed by $C_{\sf sparse+latent}$ to the graph obtained by thresholding the sample correlations.   Thresholded sample correlations have been used in several studies to infer pairwise interactions \cite{Golshani:2009, Feldt:2011, Malmersjo:2013}.  We set the threshold to match the connectivity of the $C_{\sf sparse+latent}$ graph.  The networks revealed by the two methods differed substantially. In the example site with 17.8\% connectivity, only 41.7\% of cell pairs connected in the $C_{\sf sparse+latent}$ graph were also connected in the graph of thresholded sample correlations (\figref{05}{F, H, and I}).   In particular, the thresholded sample correlations were predominantly positive whereas 43.3\% of the interactions in $C_{\sf sparse+latent}$ were negative.  Most of the negative interactions corresponded to low sample correlations (\figref{05}{F}).  Thus the $C_{\sf sparse+latent}$ estimates reveal negative interactions that account for low correlations where high correlations are predicted from all the other interactions.

The average partial correlations revealed by $C_{\sf sparse+latent}$ estimates in all 31 sites were about 5 times lower than the sample correlations and less variable across sites (\figref{06}{A}): the coefficient of variation of the average sample correlations across all sites was 0.41 whereas that of the average partial correlations was only 0.35.  The average node degree of the sparse component of the partial correlations and the number of inferred latent units varied widely between sites but generally increased with recorded population size (\figref{06}{B and C}). However, there was an inverse relationship between the number of latent units and the average node degree (\figref{06}{D}). Several sites, even with relatively large population sizes, had fairly few pairwise interactions and were dominated by latent units.  These differences may be explained by differences in brain states, recording quality (neuropil contamination, motion).

@@

We also examined the relationship of the correlation structure to the differences in orientation preference and to the geometric distances between the cell pairs (Fig.\;\ref{fig:07}).  The partial correlations of the $C_{\sf sparse+latent}$ estimates fell more rapidly with the difference in preferred orientation (\figref{07}{A}) than sample correlations. They also decreased more rapidly with geometric distance, both laterally  (\figref{07}{B}) and in depth (\figref{07}{B}).

@@

Furthermore, the positive and negative interactions in the $C_{\sf sparse+latent}$ estimator were organized differently: The probability of positive interactions fell rapidly with differences in the preferred orientation (\figref{07}{D}), lateral displacements (\figref{07}{E}), and displacement in depth (\figref{07}{F}) while the negative interactions were much less selective (\figref{07}{D--F}).
